---
title: "Geospatial Mapping"
format: pdf
editor: visual
editor_options: 
  chunk_output_type: console
---

## Libraries/Packages

```{r}
#| label: load-pkg-data
#| echo: false
#| message: false
#| warning: false
library(knitr) 
library(tidyverse)
library(patchwork)
library(stringr)
library(usmap)
library(maps)
library(janitor)
library(readxl)
library(stringr)

```

```{r}
#| label: load-data 
#| echo: false 
#| message: false

Sys.setenv("VROOM_CONNECTION_SIZE" = 131072 * 6)

test_admindist_gys <- readr::read_csv('data/EDOpportunity/seda2023_admindist_poolsub_gys_updated_20240205.csv')

income_district <- readr::read_csv("data/ACS/ACSST1Y2021.S1902-2024-04-10T232130.csv")

school_districts <- read_excel("data/sdlist-21.xls", skip = 2)
```

#### How many counties have multiple school districts?

```{r}
#| label: district-counts

county_district_count <- school_districts |>
  group_by(`State FIPS`,`County FIPS`) |>
  count() |>
  group_by(n) |>
  count()

district_count_plot <- ggplot() +
  geom_point(county_district_count, mapping = aes(x = n, y = nn))
```

```{r}
#| label: income-data-prep

#Data Cleaning to Fix Column Names and Allignment
#Only 737 school districts in this dataset, the education data has 7390 school districts

income_district_cleaned <- income_district |>
  clean_names()

income_district_estimates <- income_district_cleaned |>
  rename(
    school_district = label_grouping
  ) |>
  mutate(
    SchoolDistrict = lag(school_district, n = 2, default = NULL)
  ) |>
  filter(
    grepl("District", school_district)
  )

income_district_percentages <- income_district_cleaned |>
  rename(
    school_district = label_grouping
  ) |>
  mutate(
    school_district = lag(school_district, n = 5, default = NULL)
  ) |>
  filter(
    grepl("District", school_district)
  )
```

```{r}
#| label: ssi-percentage
#Percentage of people that receive social security income


ssi_percent_district <- income_district_percentages |>
  mutate(
    state = word(school_district, -1),
    school_district = word(school_district, start = 1, end = -4),
    state = state.abb[match(state,state.name)]
  )

write.csv(ssi_percent_district, "data/acs_income_2021.csv")


ssi_percent_joined <- ssi_percent_district |>
  left_join(
    test_admindist_gys,
    by = join_by(school_district == sedaadminname, state == stateabb)
  )


# After filtering, only 248 school districts
ssi_math <- ssi_percent_joined |>
  filter(
    subject == "mth",
    subgroup == "all"
  ) |>
  mutate(
    ol_diff_2022_2023 = gys_mn_2023_ol - gys_mn_2022_ol,
    ol_diff_2019_2022 = gys_mn_2022_ol - gys_mn_2019_ol,
  )

ssi_rla <- ssi_percent_joined |>
  filter(
    subject == "rla",
    subgroup == "all"
  ) |>
  mutate(
    ol_diff_2022_2023 = gys_mn_2023_ol - gys_mn_2022_ol,
    ol_diff_2019_2022 = gys_mn_2022_ol - gys_mn_2019_ol,
  )

ssi_math |>
  ggplot() +
  geom_point(
    aes(
      x = household_income_all_households_with_social_security_income,
      y = gys_mn_2023_ol)
    )

ssi_rla |>
  ggplot() +
  geom_point(
    aes(
      x = household_income_all_households_with_social_security_income,
      y = gys_mn_2022_ol)
    )

test_selected_years <- test_admindist_gys |>
  select(
    sedaadmin, sedaadminname, stateabb, subject, subgroup, 
    gys_mn_2019_ol, gys_mn_2022_ol, gys_mn_2023_ol
  ) 

test_wider <- test_selected_years |>
  pivot_wider(
    names_from = subject,
    values_from = c(gys_mn_2019_ol, gys_mn_2022_ol, gys_mn_2023_ol)
  )


ggplot(test_wider |>  filter(subgroup == "all")) +
  geom_point(
    aes(
      x = gys_mn_2023_ol_mth,
      y = gys_mn_2023_ol_rla,
    )
  ) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0)

write.csv(test_wider, "data/test_admindist_wider.csv")

```

Todo (4/19/2024)

-   Potentially pandemic affected survey coverage

    -   2019 may have most coverage

    -   2022 would be post pandemic

-   ACS data from one year is usually generalizable to other years

-   Shiny App Development

    -   Figure out what other variables to filter/customize by

    -   Drop-down menu variables

        -   Make all the labels human readable (not default)

        -   Make choices also affect axis + plot titles.

        -   Keep the configuration menu on the side and affect all pages

        -   Different ACS variables on the x axis

        -   Math or Reading Scores

        -   Others?

        -   Keep subgroup (demphasize)

    -   Considerations about plots

        -   Add in variable correlation

        -   Fix the y-axis scale

        -   Make it easier to distinguish above and below zero (improvement/non-improvement)

        -   For dynamic axis/plot titles

            -   Try to automate label creation (replace underscores with , capitalize every word)

        -   Subtitle: For \_\_\_ students in \_\_\_ districts

        -   Side: Consider making x-axis static within a variable

        -   Side: Interactivity with points showing popups

    -   About Page

        -   Explain the project, where the data comes from, its coverage, etc.

        -   Concise version of the report

        -   Coverage Map

            -   State map colored by count of school district

        -   Explain limitations and issues with datasets and plots

Todo (4/26/24)

App

-   Look into Expansion with other variables or datasets

-   Make all labels human readable

    -   Make them more intuitive

-   Interactivity?

-   About page

-   Use the fixed color scale for all plots going forward

-   Plots

    -   Make the points bigger

    -   Point outlining

    -   Trend Line

        -   Option to toggle it

        -   Change trend line color?

    -   Potentially plot year to year (2022 to 2023) to see more variation in Math vs Reading

-   Sidebar Menu Expansion Options

    -   Possibility to add a new drop down or exxpand the variable menu

    -   Make menus hide if they don't apply to the current tab

-   Download button on the data page

Writeup:

-   Project Report

    -   Summarizing reading and research

    -   Lit review/background

    -   Dataset descriptions

        -   Academic performance data

            -   Explain their grade calculation

        -   ACS Data

    -   Methodology

        -   Data prep + Data cleaning

        -   Consideration for district converage

    -   Results

        -   Description of the app

        -   Takeaways + Conclusions

    -   Discussion/Conclusion

        -   Overall takeways answering the research question

            -   The state of education post-pandemic

        -   LImitations of my analysis/research

        -   Future research steps

-   Reflection on the project and experience

-   Citing code/libraries

    -   Only big ones like tidyverse and shiny

Future?

-   Competitions, contests, etc.

-   Ask Mine
